Sparse Function-space Representation of Neural Networks

### 2023-09-01 workshops/icml-duality/main.tex
fixed eqs to use subequations

% \newpage
\section{\our: Sparse function-space representation of neural networks}\label{sec:sfr}
%\paragraph{GP in the dual parameters}
In this section, we present our method, named \our, which converts a trained NN into a GP.
\our is built upon a dual parameterization of the GP posterior, i.e. an alternative parameterization.
Early work by \citet{csato2002sparse} showed that Expectation Propagation gives rise to a dual parameterization.
More recent work by \citet{adam2021dual,chang2023memory} showed a different dual parameterization arising from the evidence lower bound for sparse variational GPs \citep{hensman2013gaussian,titsias2009variational}.
The dual parameterization from \citet{adam2021dual,chang2023memory}, consists of parameters $\valpha$ and $\vbeta$, which gives rise to the predictive posterior,
% , we can build on the insight from \citet{csato2002sparse} to obtain a `dual' parameterization, $\valpha$ and $\vbeta$,
% The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
% %
\begin{align}  \label{eq:gp_pred}
  \myexpect_{p(f_i \mid\vy)}[f_i] &= \vk_{\vx i}^\top \valpha, \\
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] &= k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i}, \nonumber
\end{align}
%
where $f_i = f_\vw(\vx_i)$ is the function output at input $\vx_i$,
the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$.
\cref{eq:gp_pred} states that the first two moments of the resultant posterior process (which may not be a GP), can be parameterized via the dual
parameters $\valpha, \vbeta \in \R^{N}$,
% which are the vectors
defined as, \looseness-1
%
\begin{equation}
\begin{split}
  \label{eq:dual_param}
  \alpha_i &\coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f}\log p(y_i \mid f) |_{f=f_i}], \\
  %\quad \text{and} \quad
  \beta_i &\coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f f}\log p(y_i \mid f_i) |_{f=f_i}].
  \end{split}
\end{equation}
%
The relationships specified are valid for generic likelihoods and involve no approximations since the expectation is under the exact posterior, but given that the model can be expressed in a kernel formulation. \cref{eq:gp_pred} and \cref{eq:dual_param} highlight that the approximate inference technique, usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.

@@

\paragraph{Dual parameters from NN}
Given that we use a Laplace approximation of the NN, we remove the expectation over the posterior \citep[see Ch.~3.4.1 in][for derivation]{rasmussen2006gaussian} and our dual parameters are given by,
%
\begin{equation}
\begin{split}
  \label{eq:dual_param_laplace}
  \hat{\alpha}_i &\coloneqq \nabla_{f}\log p(y_i \mid f) |_{f=f_i} , \\
  %\quad \text{and} \quad
  \hat{\beta}_i &\coloneqq - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i}.
  \end{split}
\end{equation}
%
Substituting \cref{eq:dual_param_laplace} into \cref{eq:gp_pred}, we obtain our GP based on the converged NN. The problem with forming predictions using \cref{eq:gp_pred} is that they cost $O(N^3)$, which limits a full GP model to application on large data sets.

@@

As demonstrated in \citet{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper exploit the dual parameters for approximate inference but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients.
We project the dual parameters onto the inducing points giving sparse dual parameters,
% In order to link the dual parameters defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual parameters,
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i=1}^N  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \MBeta_{\vu} =  \sum_{i=1}^N \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual parameters are now a sum over \emph{all data points}, with $\valpha_{\vu} \in \R^{M}$ and $\MBeta_{\vu} \in \R^{M  \times M}$. Using this sparse definition of the dual parameters, our sparse GP posterior takes the following form:
\begin{align}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] &= \vkzs^{\T} \MKzz^{-1} \valpha_{\vu}, \\
   %\quad \text{and} \quad
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  &= k_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \MBeta_{\vu})^{-1} ]\vkzs, \nonumber
\end{align}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\{\vz_j\}_{i=1}^M$, $\vz \in \R^{D}$, instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged NN are $(\hat{\alpha}_i, \hat{\beta}_i)$ (\cref{eq:dual_param_laplace}) and a kernel function $\kappa$ (\cref{eq-laplace-approx-function-space}). Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$, with $M \ll N$.  Crucially, given the structure of our probabilistic model, our sparse dual parameters \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

>>>>

% \newpage
\section{\our: Sparse function-space representation of neural networks}\label{sec:sfr}
%\paragraph{GP in the dual parameters}
In this section, we present our method, named \our, which converts a trained NN into a GP.
\our is built upon a dual parameterization of the GP posterior, i.e. an alternative parameterization.
Early work by \citet{csato2002sparse} showed that Expectation Propagation gives rise to a dual parameterization.
More recent work by \citet{adam2021dual,chang2023memory} showed a different dual parameterization arising from the evidence lower bound for sparse variational GPs \citep{hensman2013gaussian,titsias2009variational}.
The dual parameterization from \citet{adam2021dual,chang2023memory}, consists of parameters $\valpha$ and $\vbeta$, which gives rise to the predictive posterior,
% , we can build on the insight from \citet{csato2002sparse} to obtain a `dual' parameterization, $\valpha$ and $\vbeta$,
% The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
% %
\begin{subequations}  \label{eq:gp_pred}
\begin{align}
  \myexpect_{p(f_i \mid\vy)}[f_i] &= \vk_{\vx i}^\top \valpha, \\
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] &= k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},  \nonumber
\end{align}
\end{subequations}
%
where $f_i = f_\vw(\vx_i)$ is the function output at input $\vx_i$,
the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$.
\cref{eq:gp_pred} states that the first two moments of the resultant posterior process (which may not be a GP), can be parameterized via the dual
parameters $\valpha, \vbeta \in \R^{N}$,
% which are the vectors
defined as, \looseness-1
%
\begin{subequations}
\label{eq:dual_param}
\begin{align}
  \alpha_i &\coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f}\log p(y_i \mid f) |_{f=f_i}], \\
  %\quad \text{and} \quad
  \beta_i &\coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f f}\log p(y_i \mid f_i) |_{f=f_i}].
  \end{align}
\end{subequations}
%
The relationships specified are valid for generic likelihoods and involve no approximations since the expectation is under the exact posterior, but given that the model can be expressed in a kernel formulation. \cref{eq:gp_pred} and \cref{eq:dual_param} highlight that the approximate inference technique, usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.

@@

\paragraph{Dual parameters from NN}
Given that we use a Laplace approximation of the NN, we remove the expectation over the posterior \citep[see Ch.~3.4.1 in][for derivation]{rasmussen2006gaussian} and our dual parameters are given by,
%
\begin{subequations}
\label{eq:dual_param_laplace}
\begin{align}
  \hat{\alpha}_i &\coloneqq \nabla_{f}\log p(y_i \mid f) |_{f=f_i} , \\
  %\quad \text{and} \quad
  \hat{\beta}_i &\coloneqq - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i}.
  \end{align}
\end{subequations}
%
Substituting \cref{eq:dual_param_laplace} into \cref{eq:gp_pred}, we obtain our GP based on the converged NN. The problem with forming predictions using \cref{eq:gp_pred} is that they cost $O(N^3)$, which limits a full GP model to application on large data sets.

@@

As demonstrated in \citet{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper exploit the dual parameters for approximate inference but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients.
We project the dual parameters onto the inducing points giving sparse dual parameters,
% In order to link the dual parameters defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual parameters,
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i=1}^N  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \MBeta_{\vu} =  \sum_{i=1}^N \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual parameters are now a sum over \emph{all data points}, with $\valpha_{\vu} \in \R^{M}$ and $\MBeta_{\vu} \in \R^{M  \times M}$. Using this sparse definition of the dual parameters, our sparse GP posterior takes the following form:
\begin{align}\label{eq:dual_sparse_post}
\begin{subequations} \label{eq:dual_sparse_post}
\begin{align}
   \myexpect_{q_{\vu}(\vf)}[f_i] &= \vkzs^{\T} \MKzz^{-1} \valpha_{\vu}, \\
   %\quad \text{and} \quad
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  &= k_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \MBeta_{\vu})^{-1} ]\vkzs, \nonumber
\end{align}
\end{subequations}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\{\vz_j\}_{i=1}^M$, $\vz \in \R^{D}$, instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged NN are $(\hat{\alpha}_i, \hat{\beta}_i)$ (\cref{eq:dual_param_laplace}) and a kernel function $\kappa$ (\cref{eq-laplace-approx-function-space}). Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$, with $M \ll N$.  Crucially, given the structure of our probabilistic model, our sparse dual parameters \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.
